---
title: "01.00-trimming-fastp-fastqc-multiqc"
author: "Sam White"
date: "2025-04-17"
output: 
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
---

# BACKGROUND

This Rmd file trims FastQ files using [fastp](https://github.com/OpenGene/fastp) [@chen2023], followed by quality checks with [FastQC](https://github.com/s-andrews/FastQC) and [MultiQC](https://multiqc.info/) [@ewels2016].

::: callout-note
If you need to download the raw sequencing reads, please see [00.00-fastqc-multiqc-raw-reads.Rmd](https://github.com/RobertsLab/project-eDNA-yellow/blob/6e9534bfbc3ebccf20f550b428bf5c0a08356c04/code/00.00-fastqc-multiqc-raw-reads.Rmdd)
:::

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(reticulate)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

Inputs:

-   Paired-end FastQs

    -   Example filename format: `Yell_July18_Site1f_S92_L001_R[12]_001.fastq.gz`

Outputs:

-   Trimmed FastQs.

    -   Example filename format: `Yell_July18_Site1f_S92_L001_R1.fastp-trim.fq.gz`

-   fastp JSON files.

-   FastQC HTML files.

-   MultiQC HTML report.

Software requirements:

-   [fastp](https://github.com/OpenGene/fastp)

-   [FastQC](https://github.com/s-andrews/FastQC)

-   [MultiQC](https://github.com/MultiQC/MultiQC) [@ewels2016]

# Set R variables

```{r R-variables, eval=TRUE}

# Data directories
data_dir <- "../data/"
raw_reads_dir <- "../data/raw-fastqs"
trimmed_fastqs_dir <- "../output/01.00-trimming-fastp-fastqc-multiqc" # Define trimmed fastqs directory

# PROGRAMS
fastp <- "/home/sam/programs/fastp"
fastqc <- "/home/sam/programs/FastQC-0.12.1/fastqc"
multiqc <- "/home/sam/programs/miniforge3/bin/multiqc"

# CPUs
threads <- 4

# Define file patterns
R1_fastq_pattern <- "*_R1_001.fastq.gz"
R2_fastq_pattern <- "*_R2_001.fastq.gz"

# Export these as environment variables for bash chunks.
Sys.setenv(
  fastp = fastp,
  fastqc = fastqc,
  data_dir = data_dir,
  multiqc = multiqc,
  raw_reads_dir = raw_reads_dir,
  threads = threads,
  R1_fastq_pattern = R1_fastq_pattern,
  R2_fastq_pattern = R2_fastq_pattern,
  trimmed_fastqs_dir = trimmed_fastqs_dir
)
```

# Fastp Trimming

[fastp](https://github.com/OpenGene/fastp) [@chen2023] is set to auto-detect Illumina adapters, as well as trim the first 10bp from each read, as past experience shows these first 10bp are more inconsistent than the remainder of the read length.

```{r fastp-trimming, engine='bash', eval=TRUE}

# Make output directories, if it doesn't exist
mkdir --parents "${trimmed_fastqs_dir}"

# Change to raw reads directory
cd "${raw_reads_dir}"

# Create arrays of fastq R1 files and sample names
for fastq in ${R1_fastq_pattern}
do
  fastq_array_R1+=("${fastq}")
  R1_names_array+=("$(echo "${fastq}" | awk -F"_" '{print $1}')")
done

# Create array of fastq R2 files
for fastq in ${R2_fastq_pattern}
do
  fastq_array_R2+=("${fastq}")
  R2_names_array+=("$(echo "${fastq}" | awk -F"_" '{print $1}')")
done

# Create list of fastq files used in analysis
# Create MD5 checksum for reference
if [ ! -f "${output_dir_top}"/raw-fastq-checksums.md5 ]; then
  for fastq in *.gz
    do
      md5sum "${fastq}" >>"${output_dir_top}"/raw-fastq-checksums.md5
  done
fi

# Run fastp on files
# Adds JSON report output for downstream usage by MultiQC
for index in "${!fastq_array_R1[@]}"
do
  R1_sample_name=$(echo "${R1_names_array[index]}")
  R2_sample_name=$(echo "${R2_names_array[index]}")
  echo "${R1_sample_name}"
  echo ""
  ${fastp} \
  --in1 ${fastq_array_R1[index]} \
  --in2 ${fastq_array_R2[index]} \
  --detect_adapter_for_pe \
  --trim_front1 10 \
  --trim_front2 10 \
  --trim_poly_g \
  --thread ${threads} \
  --html "${trimmed_fastqs_dir}"/"${R1_sample_name}".fastp-trim.report.html \
  --json "${trimmed_fastqs_dir}"/"${R1_sample_name}".fastp-trim.report.json \
  --out1 "${trimmed_fastqs_dir}"/"${R1_sample_name}"_R1.fastp-trim.fq.gz \
  --out2 "${trimmed_fastqs_dir}"/"${R2_sample_name}"_R2.fastp-trim.fq.gz \
  2>> "${trimmed_fastqs_dir}"/fastp.stderr

  # Generate md5 checksums for newly trimmed files
  cd "${trimmed_fastqs_dir}"
  md5sum "${R1_sample_name}"_R1.fastp-trim.fq.gz > "${R1_sample_name}"_R1.fastp-trim.fq.gz.md5
  md5sum "${R2_sample_name}"_R2.fastp-trim.fq.gz > "${R2_sample_name}"_R2.fastp-trim.fq.gz.md5
  
  # Change back to previous directory
  cd -
done
```
